A full approximation scheme multilevel method \\ for nonlinear variational inequalities

### 2023-12-29 paper/fascd.tex
remove the Figure I just put in, and a few more small edits

Prior multilevel algorithms for VIs have significant limitations compared to the FASCD framework.  Listed chronologically, here are some comparisons:
\begin{itemize}
\item The projected full approximation scheme (PFAS) algorithm of Brandt \& Cryer \cite{BrandtCryer1983} applies to linear complementarity problems and it uses projected (truncated) Gauss--Seidel smoothers.  While our method reduces to general, nonlinear FAS for PDEs when the inequality constraints are removed, PFAS reduces to linear multigrid.
\item The multilevel CD iteration of Tai \cite{Tai2003} is formulated in terms of an objective function; as stated it only applies to optimization-type problems.  Furthermore, each of its coarser-level problems refers to the finest-level discretization.
\item The last-mentioned limitation of the \cite{Tai2003} method is removed by Gr\"aser \& Kornhuber's \cite[Algorithm 4.2]{GraeserKornhuber2009} reformulation using defect constraints and monotone injection operators.  We adopt these techniques; see Section \ref{sec:cdmultilevel} below.  However, the monotone multigrid methods of Kornhuber and co-authors \cite{GraeserKornhuber2009,Kornhuber1994} appear to be formulated only with Gauss--Seidel type smoothers, and may have been applied only to optimization-type VI problems \cite[for example]{JouvetGraeser2013}.  Furthermore, unlike the fastest variants of monotone multigrid, no zeroing of the basis functions associated with the active set (truncation) is employed in FASCD, and this simplifies the implementation.  FASCD coarse grid corrections can indeed update the finest-level active set, and the numerical evidence indicates that FASCD coarse corrections are effective even when the coarse meshes are far too coarse to resolve the active set.  On a problem with a complicated coincidence set, for which truncated Newton/monotone multigrid methods were apparently designed \cite[problem 7.1.1]{GraeserKornhuber2009}, the performance of FASCD is at least as good (Example \ref{ex:results:classical}).
\item Another work which combines FAS and VIs is \cite{HintermullerKopacka2011}, but the problems solved, and the approach, are completely different.  For an optimal control problem with an elliptic VI constraint, \cite{HintermullerKopacka2011} applies FAS to a system of PDEs derived from penalizing the VI and the control constraints; their FAS algorithm is not applied to the VI itself while ours is.
\end{itemize}

@@

\begin{example} \label{ex:results:classical}
Consider the classical unilateral 2D Laplacian obstacle problem, namely Example \ref{ex:plaplacian} with $p=2$ and admissible set $\mathcal{K} = \{v \ge \psi\}$.  We consider two cases, over square domains $\Omega$, with different coincidence (active) sets $\{x\in\Omega \,:\, u(x)=\psi(x)\}$ (Figure \ref{fig:results:classical}).  The standard ``ball'' problem has the unit upper hemisphere as the obstacle \cite[Chapter 12]{Bueler2021}.  The ``spiral'' problem's obstacle is constructed to produce the coincidence set shown \cite[problem 7.1.1]{GraeserKornhuber2009}.  For both cases we used a regular coarsest mesh with $m_0=41$ nodes (64 triangles), and then the finest mesh had $m_{10} \approx 3.4 \times 10^7$ nodes after 10 refinements.  A single RS Newton iteration with three IC-preconditioned CG iterations was used as the smoother, and the initial iterate was $w=\max\{0,\psi\}$.

@@

@@

%REGENERATE using reproduce/table2.py at git@bitbucket.org:pefarrell/fascd.git
\begin{figure}[ht]
\centering
\includegraphics[width=0.65\textwidth]{asymprates.pdf}
\caption{V-cycle asymptotic convergence rates in Example \ref{ex:results:classical}.}
\label{fig:results:asymp}
\end{figure}

@@

Iteration results in Table \ref{tab:results:fastplap1d} show a strong contrast between $V(1,0)$ and $V(0,1)$ results, chosen by setting \pr{up} $=0$ and \pr{down} $=0$ in Algorithm \ref{alg:fascd}, respectively.  The $V(1,0)$ cycles, essentially Algorithm 4.2 in \cite{GraeserKornhuber2009} but with a stronger smoother, often do not converge.  By contrast, $V(0,1)$ cycles, which up-smooth in the larger sets $\mathcal{U}^j$ (Section \ref{sec:cdmultilevel}), are as good as the default $V(1,1)$ cycles.  FMG is even more efficient; a single iteration suffices at all resolutions.   Note that convergence is at rate $O(h^{2.0})$; compare \cite{ChoeLewis1991}.
\end{example}

@@

% see git@bitbucket.org:pefarrell/fascd.git: examples/sia.py
% and reproduce/table5.sh
\begin{table}[ht]
\centering
\begin{tabular}{c@{\hskip 4mm}c@{\hskip 4mm}c@{\hskip 7mm}c@{\hskip 4mm}cc}
\toprule
$P$ & \emph{levels} & $m_J$ & \emph{iterations} & \emph{time} (s) \\
\midrule
 1 & 6 & $641^2$ & +2 & 98 \\
 4 & 7 & $1281^2$ & +2 & 98 \\
 16 & 8 & $2561^2$ & +2 & 124 \\
 64 & 9 & $5121^2$ & +2 & 136 \\
 256 & 10 & $10241^2$ & +1 & 95 \\
 1024 & 11 & $20481^2$ & +1 & 177 \\
 \bottomrule
\end{tabular}
\bigskip
\caption{Parallel weak scaling FMG cycle results (V-cycle iterations to satisfy \eqref{eq:stoppingcriterion}, and wall clock time) for Example \ref{ex:results:sia}.}
\label{tab:results:siaweak}
\end{table}
\end{example}

>>>>

Prior multilevel algorithms for VIs have significant limitations compared to the FASCD framework.  Listed chronologically, here are some comparisons:
\begin{itemize}
\item The projected full approximation scheme (PFAS) algorithm of Brandt \& Cryer \cite{BrandtCryer1983} applies to linear complementarity problems and it uses projected (truncated) Gauss--Seidel smoothers.  While our method reduces to general, nonlinear FAS for PDEs when the inequality constraints are removed, PFAS reduces to linear multigrid.
\item The multilevel CD iteration of Tai \cite{Tai2003} is formulated in terms of an objective function; as stated it only applies to optimization-type problems.  Furthermore, each of its coarser-level problems refers to the finest-level discretization.  FASCD is both more general, as it applies to non-variational monotone operators (Section \ref{sec:vi}), and more efficient on each coarser level.
\item The last-mentioned limitation of the \cite{Tai2003} method is removed by Gr\"aser \& Kornhuber's \cite[Algorithm 4.2]{GraeserKornhuber2009} reformulation using defect constraints and monotone injection operators.  We adopt these techniques; see Section \ref{sec:cdmultilevel} below.  However, the monotone multigrid methods of Kornhuber and co-authors \cite{GraeserKornhuber2009,Kornhuber1994} appear to be formulated only with Gauss--Seidel type smoothers, and may have been applied only to optimization-type VI problems \cite[for example]{JouvetGraeser2013}.  Furthermore, unlike the fastest variants of monotone multigrid, no zeroing of the basis functions associated with the active set (truncation) is employed in FASCD, and this simplifies the implementation.  FASCD coarse grid corrections can indeed update the finest-level active set, and the numerical evidence indicates that FASCD coarse corrections are effective even when the coarse meshes are far too coarse to resolve the active set.  On a problem with a complicated coincidence set, for which truncated Newton/monotone multigrid methods were apparently designed \cite[problem 7.1.1]{GraeserKornhuber2009}, the performance of FASCD is at least as good (Example \ref{ex:results:classical}).
\item Another work which combines FAS and VIs is \cite{HintermullerKopacka2011}, but the problems solved, and the approach, are completely different.  For an optimal control problem with an elliptic VI constraint, \cite{HintermullerKopacka2011} applies FAS to a system of PDEs derived from penalizing the VI and the control constraints; their FAS algorithm is not applied to the VI itself while ours is.
\end{itemize}

@@

\begin{example} \label{ex:results:classical}
Consider the classical unilateral 2D Laplacian obstacle problem, namely Example \ref{ex:plaplacian} with $p=2$ and admissible set $\mathcal{K} = \{v \ge \psi\}$.  We consider two cases, over square domains $\Omega$, with different coincidence (active) sets $\{x\in\Omega \,:\, u(x)=\psi(x)\}$ (Figure \ref{fig:results:classical}).  The standard ``ball'' problem has the unit upper hemisphere as the obstacle \cite[Chapter 12]{Bueler2021}.  The ``spiral'' problem's obstacle is constructed to produce the coincidence set shown \cite[problem 7.1.1]{GraeserKornhuber2009}.  For both cases we used $P_1$ elements on a regular coarsest mesh with $m_0=41$ nodes (64 triangles), and then the finest mesh had $m_{10} \approx 3.4 \times 10^7$ nodes after 10 refinements.  A single RS Newton iteration with three IC-preconditioned CG iterations was used as the smoother, and the initial iterate was $w=\max\{0,\psi\}$.

@@

The ball problem has a known exact solution $u\in W^{2,\infty}(\Omega)$ \cite{Bueler2021,Tai2003}.  Our measured convergence rate is $\|u^\star-u^J\|_{L^2(\Omega)}=O(h^{1.98})$, as expected for $P_1$ elements.  FIXME: CITE CIARLET OR GLOWINSKI

@@

%REGENERATE using reproduce/table2.py at git@bitbucket.org:pefarrell/fascd.git
\begin{figure}[ht]
\centering
\includegraphics[width=0.65\textwidth]{asymprates.pdf}
\caption{V-cycle asymptotic rates in Example \ref{ex:results:classical}.}
\label{fig:results:asymp}
\end{figure}

@@

Iteration results in Table \ref{tab:results:fastplap1d} show a strong contrast between $V(1,0)$ and $V(0,1)$ results, chosen by setting \pr{up} $=0$ and \pr{down} $=0$ in Algorithm \ref{alg:fascd}, respectively.  The $V(1,0)$ cycles, essentially Algorithm 4.2 in \cite{GraeserKornhuber2009} but with a stronger smoother, often do not converge.  By contrast, $V(0,1)$ cycles, which up-smooth in the larger sets $\mathcal{U}^j$ (Section \ref{sec:cdmultilevel}), are as good as the default $V(1,1)$ cycles.  FMG is even more efficient; a single iteration suffices at all resolutions.   Note that measured $L^\infty(\Omega)$ convergence is $O(h^{2.00})$; compare \cite{ChoeLewis1991}.
\end{example}

@@

% see git@bitbucket.org:pefarrell/fascd.git: examples/sia.py
% and reproduce/table5.sh
\begin{table}[ht]
\centering
\begin{tabular}{c@{\hskip 4mm}c@{\hskip 4mm}c@{\hskip 7mm}c@{\hskip 4mm}cc}
\toprule
$P$ & \emph{levels} & $m_J$ & \emph{iterations} & \emph{time} (s) \\
\midrule
 1 & 6 & $641^2$ & +2 & 98 \\
 4 & 7 & $1281^2$ & +2 & 98 \\
 16 & 8 & $2561^2$ & +2 & 124 \\
 64 & 9 & $5121^2$ & +2 & 136 \\
 256 & 10 & $10241^2$ & +1 & 95 \\
 1024 & 11 & $20481^2$ & +1 & 177 \\
 \bottomrule
\end{tabular}
\bigskip
\caption{Parallel weak scaling FMG cycle results (additional V-cycle iterations to satisfy \eqref{eq:stoppingcriterion}, and wall clock time) for Example \ref{ex:results:sia}.}
\label{tab:results:siaweak}
\end{table}
\end{example}
